A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd

The structure of population is a key variable on the incidence. Several papers related the relationship between a fast trasnmision of COVID-19 and population density (e.g. cumulative incidence in urban areas like NYC) and therefore a positive sign will be expected. Note that our analysis is at provincial level - not a urban level - and therefore is necessary to rethinking the relationship. In Spain the population is concentrate (with some exception) in the mediterranean coast (@Goerlich2013population), in general with high temperature and high humidity. So, positive or negative sign will be expected in the model.

All those variables will we include in log to capture non-linar realtionship.

Transit.

@@

```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people (map omits Islas Canarias)", warning=FALSE, cache = TRUE}
# Retrieve the coordinates of Madrid
madrid_xy <-  st_centroid(provinces_spain %>% filter(province == "Madrid")) %>% st_coordinates()

@@

```{r map-controls, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:map-controls}Spatial distribution of control variables by province (maps omits Islas Canarias)", cache = TRUE}
# GDP per capita:
q <- quantile(provinces_spain$GDPpc)
gdppc.plot <- provinces_spain %>%
  mutate(GDPpc = GDPpc,
         GDPpc = case_when(GDPpc < q[2] ~ "1st",
                             GDPpc > q[2] & GDPpc <= q[3]  ~ "2nd",
                             GDPpc > q[3] & GDPpc <= q[4] ~ "3rd",
                             GDPpc > q[4] ~ "4th")) %>%
  #filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = GDPpc)) +
  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom") +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 8, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

@@

```{r goodness-of-fit, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:goodness-of-fit} Goodness of fit of the SUR systems: by date and pooled"}
data.frame(Model1 = sur.slm_lag8$R2, 
           Model2 = sur.slm_lag11$R2,
           Model3 = sur.slm_lag11w$R2) %>%
  slice(2:n()) %>%
  rownames_to_column(var = "Equation") %>%
  mutate(Date = seq(ymd("2020-03-13"), 
                    ymd("2020-04-11"), 
                    by = "days")) %>%
  pivot_longer(cols = starts_with("M"), names_to = "Model", values_to = "R2") %>%
  mutate(Model = factor(Model, 
                        levels = c("Model1", "Model2", "Model3"),
                        labels = c("Model 1", "Model 2", "Model 3"))) %>%
  ggplot(aes(x = Date, y = R2, color = Model, shape = Model)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Model 1" = "blue", "Model 2" = "orange", "Model 3" = "green") ) + 
  annotate(geom = "text", 
           label = c(paste("Model 1: pooled-R2=", 
                           round(sur.slm_lag8$R2[1], 4)),
                     paste("Model 2: pooled-R2=", 
                           round(sur.slm_lag11$R2[1], 4)),
                     paste("Model 3: pooled-R2=", 
                           round(sur.slm_lag11w$R2[1], 4))),
           x = as_date("2020-04-01"),
           y = c(0.3, 0.27, 0.24),
           family = "serif",
           size = 3.5) +
  theme_tufte()

@@

```{r impacts}
Wsur <- as(spdep::listw2mat(listw), "CsparseMatrix")
trsur <- spatialreg::trW(Wsur, type = "MC")
impacts.sur.slm_lag11 <- impactspsur(sur.slm_lag11, tr = trsur,R = 1000)

>>>>

The third population variable the we consider is population density, since it directly affects the contact patterns and contact rates between individuals in a population [@Hu2013scaling]. The evidence available suggests a positive relationship between the transmission of COVID-19 and population density (e.g. cumulative incidence in urban areas like NYC). For this reason, we anticipate a positive relationship between population density and the incidence of the disease.

The last control variable is the presence of mass transit systems in a province. Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility, namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches [@Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing. After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.

@@

```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people", warning=FALSE, cache = TRUE}
# Retrieve the coordinates of Madrid
madrid_xy <-  st_centroid(provinces_spain %>% filter(province == "Madrid")) %>% st_coordinates()

@@

```{r map-controls, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:map-controls}Spatial distribution of control variables by province", warning=FALSE, cache = TRUE}
# GDP per capita:
q <- quantile(provinces_spain$GDPpc)
gdppc.plot <- provinces_spain %>%
  mutate(GDPpc = GDPpc,
         GDPpc = case_when(GDPpc < q[2] ~ "1st",
                             GDPpc > q[2] & GDPpc <= q[3]  ~ "2nd",
                             GDPpc > q[3] & GDPpc <= q[4] ~ "3rd",
                             GDPpc > q[4] ~ "4th")) %>%
  #filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = GDPpc)) +
  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom") +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 8, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

@@

```{r goodness-of-fit, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:goodness-of-fit} Goodness of fit of the SUR systems: by date and pooled"}
data.frame(Model1 = sur.slm_lag8$R2, 
           Model2 = sur.slm_lag11$R2,
           Model3 = sur.slm_lag11w$R2) %>%
  slice(2:n()) %>%
  rownames_to_column(var = "Equation") %>%
  mutate(Date = seq(ymd("2020-03-13"), 
                    ymd("2020-04-11"), 
                    by = "days")) %>%
  pivot_longer(cols = starts_with("M"), names_to = "Model", values_to = "R2") %>%
  mutate(Model = factor(Model, 
                        levels = c("Model1", "Model2", "Model3"),
                        labels = c("Model 1: lag8", "Model 2: lag11", "Model 3: lag11w"))) %>%
  ggplot(aes(x = Date, y = R2, color = Model, shape = Model)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Model 1: lag8" = "blue", "Model 2: lag11" = "orange", "Model 3: lag11w" = "green") ) + 
  annotate(geom = "text", 
           label = c(paste("pooled-R2 (Model 1: lag8)=", 
                           round(sur.slm_lag8$R2[1], 4)),
                     paste("pooled-R2 (Model 2: lag11)=", 
                           round(sur.slm_lag11$R2[1], 4)),
                     paste("pooled-R2 (Model 3: lag11w)=", 
                           round(sur.slm_lag11w$R2[1], 4))),
           x = as_date("2020-04-01"),
           y = c(0.3, 0.27, 0.24),
           family = "serif",
           size = 3.5) +
  theme_tufte()

@@

```{r impacts, echo=FALSE}
Wsur <- as(spdep::listw2mat(listw), "CsparseMatrix")
trsur <- spatialreg::trW(Wsur, type = "MC")
impacts.sur.slm_lag11 <- impactspsur(sur.slm_lag11, tr = trsur,R = 1000)
